\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2021-01-05 manuscript.Rnw
Manuscript updates

Analyses were performed using \pkg{hesim} version \code{0.5.0}.

@@

This model can be simulated using either of the model types described in Section~\ref{sec:model-taxonomy}. We demonstrate use of all three below, starting with the iCTSTM.

@@

The target population, treatment strategies, and (non-death) health states are stored in a \code{hesim_data} object. There are 3 treatment strategies: standard of care (SOC) and two new interventions (New 1 and New 2). The target population consists of a heterogeneous population of 1,000 patients, which is large enough to ensure that averages across patients in the individual simulation are reasonably stable. As we will see in the next section, patients can also be placed into subgroups with a \code{grp_id} column in the \code{patients} table; if left unspecified as done here, then all patients are assumed to belong to a single subgroup.

@@

@@

For a given subgroup, QALYs and costs are simulated for each PSA sample, treatment strategy, and health state. They are summarized with the \code{$summarize()} method, which first sums across health states and then computes a (weighted) average across patients, by PSA sample, treatment strategy, and subgroup. The output is a generic cost-effectiveness object (of class \code{ce}) that can be used to perform CEA as demonstrated in Section~\ref{sec:cea}. Here, we use the \code{summary()} method to summarize the results and make use the ``pipes'' from the \pkg{magrittr} to nicely format the output.

@@

@@

We first simulate survival curves by month for 20 years. State probabilities are easily computed from these survival curves with the \code{sim_stateprobs()} method. The area under the PFS curve is the proportion of patients in stable disease while the area between the PFS and OS curves is the proportion of patient whose disease has progressed.

<<psm_sim_survival>>=
times <- seq(0, 20, by = .1)
psm$sim_survival(t = times)
head(psm$survival_)
@

@@

QALYs and costs are simulated using the same \code{sim_qalys()} and \code{sim_costs()} methods as in the previous section, and the results are summarized in the same fashion as well.

@@

In Sections~\ref{sec:ictstm-example} and \ref{sec:psm-example}, we constructed the disease model using parameter (i.e., \code{params_surv}) objects. Here, we will use a transformed parameter (i.e., \code{tparams_transprobs}) object, which is a very flexible way to construct the transition probability matrices for a cDTSTM. A \code{tparams_transprobs} object requires a 3-dimensional array of transition matrices, sorted by PSA sample, treatment strategy, patient, and optionally (in a time-inhomogeneous model) time interval.

@@

Identifiers for each matrix in the array can then be generated with \code{tpmatrix_id()}.

@@

<<rr>>=
x_rr <- as.matrix(transmod_data[, .(soc, new1, new2)])
rr <- cbind(xbeta(x_rr, params_rr_rng$rr_12),
            xbeta(x_rr, params_rr_rng$rr_13))
@

@@

and the results are summarized.

@@

\subsubsection{Incremental cost-effectiveness ratio}
The \code{cea_pw()} function computes the ICER. A summary table can be produced with \code{icer()} and \code{format()} can format the table for pretty printing. The argument \code{k} is the WTP threshold and is used to compute the INMB. Estimates of incremental QALYs and incremental costs are computed by averaging over PSA samples. By default, 95\% confidence intervals are presented and are computed using quantiles from the PSA.

@@

\subsubsection{Representing decision uncertainty}
A number of common measures can be used to represent decision uncertainty from a CEA, including cost-effectiveness planes, cost-effectiveness acceptability curves (CEAC), cost-effectiveness acceptability frontiers (CEAF), and the expected value of perfect information (EVPI). Plots can be quickly generated with \code{plot_ceplane()}, \code{plot_ceac()}, \code{plot_ceaf()}, and \code{plot_evpi()} functions, respectively. Example plots from the code that follows are shown in Figure~\ref{fig:cea-plots}

@@

One drawback of the CEAC is that the probability of being cost-effective cannot be used to determine the optimal treatment option. Instead, if the objective is to maximize health gain, then decisions should be based on the expected NMB. The CEAF, which plots the probability that the optimal treatment strategy (i.e., the strategy with the highest expected NMB) is cost-effective, is appropriate in this context.

A CEAF curve can be created by using the \code{best} column from the \code{mce} element to subset to the treatment strategy with the highest expected NMB for each WTP value.

@@

Alternative approaches might be appropriate in other contexts. For example, if evenly spaced panel data is available, a multinomial logistic regressions can be used to predict transition probabilities. On the other hand, some users may prefer a functional approach that is arguably more human readable. To facilitate this type of analysis, a \code{define_model()} block can be used to construct transition probability matrices using nonstandard evaluation. Each possible combination of the parameters and input data (treatment strategies and patients) is automatically constructed and the user can write custom expressions that create the transition probability matrices as a function of the parameters and input data. The same \code{define_model()} block may also be used to construct utility and cost models, if desired. Example that leverage multinomial logistic regression and \code{define_model()} are provided on the package website. The latter is used to reproduce two examples (a simple Markov cohort model of an HIV treatment and a time-inhomogeneous Markov cohort model of total hip replacement) from the reference textbook by \citet{briggs2006decision}.

@@

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models. It supports prediction of state probabilities and simulation of multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health-economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is also computationally inefficient for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions where efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran 100 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. The model completed in 0.44 second in \pkg{hesim} and took 34 minutes in \pkg{mstate}, showing that computational considerations are not just an academic concern.

@@

%% -- Conclusion ---------------------------------------------------------------
\section{Conclusion} \label{sec:conclusion}
\pkg{hesim} is a modular and computationally efficient \proglang{R} package for development and analysis of simulation models for economic evaluation. Discrete time cohort and continuous time individual-level models are supported, encompassing Markov (time-homogeneous and time-inhomogeneous) and semi-Markov processes as well as partitioned survival analysis. To facilitate integration of statistical methods and economic models and use of both patient- and aggregate-level data, parameters can be estimated by fitting a statistical model with \proglang{R}, inputting parameter estimates from an external source, or a combination of the two. Simulated costs and QALYs from a PSA can be used for for decision-analysis within a cost-effectiveness framework, allowing users to represent decision uncertainty (e.g., cost-effectiveness planes, cost-effectiveness acceptability curves, and cost-effectiveness acceptability frontiers), and conduct value of information analysis. Algorithms are efficiently coded and written in \proglang{C++} so that approaches that have been considered computationally intensive such as individual-level simulations and PSA can run very quickly. The modular design makes modeling more flexible and it straightforward to add new features.

>>>>

Analyses were performed using \pkg{hesim} version \code{0.5.0}. All code including a replication \proglang{R} script for this paper is available at \url{https://github.com/hesim-dev/hesim-manuscript}.

@@

This model can be simulated using any of the model types described in Section~\ref{sec:model-taxonomy}. We demonstrate use of all three below, starting with the iCTSTM.

@@

The target population, treatment strategies, and (non-death) health states are stored in a \code{hesim_data} object and identified with integer valued identification variables (\code{patient_id}, \code{strategy_id}, and \code{state_id}). There are 3 treatment strategies: standard of care (SOC) and two new interventions (New 1 and New 2). The target population consists of a heterogeneous population of 1,000 patients, which is large enough to ensure that averages across patients in the individual simulation are reasonably stable. As we will see in the next section, patients can also be placed into subgroups (with a \code{grp_id} column) or given weights (with a \code{patient_wt} column) in the \code{patients} table; if left unspecified as done here, then all patients are assumed to belong to a single subgroup and are weighted equally.

@@

Labels for the identification variables in a \code{hesim_data} object can be generated with \code{get_labels()}. These can (as will be shown below), in turn, be passed to plotting and summary functions to create tables and figures with more informative labels.

@@

For a given subgroup, QALYs and costs are simulated for each PSA sample, treatment strategy, and health state. They are summarized with the \code{$summarize()} method, which first sums across health states and then computes a (weighted) average across patients, by PSA sample, treatment strategy, and subgroup. The output is a generic cost-effectiveness object (of class \code{ce}) that can be used to perform CEA as demonstrated in Section~\ref{sec:cea}. Here, we use the \code{summary()} method to summarize the results and make use the ``pipes'' from the \pkg{magrittr} to nicely format the output. Labels for the treatment strategies are based on output of \code{get_labels()} produced above.

@@

Since there are now subgroups, we generate new labels that include the subgroup variables.

@@

We first simulate survival curves by month for 30 years. State probabilities are easily computed from these survival curves with the \code{sim_stateprobs()} method. The area under the PFS curve is the proportion of patients in stable disease while the area between the PFS and OS curves is the proportion of patient whose disease has progressed.

<<psm_sim_survival>>=
times <- seq(0, 30, by = .1)
psm$sim_survival(t = times)
head(psm$survival_)
@

@@

QALYs and costs are simulated using the same \code{sim_qalys()} and \code{sim_costs()} methods as in the previous section, and the results are summarized in the same fashion as well. A difference, however, is that we now use the \code{by_grp = TRUE} option to summarize costs and QALYs by subgroup (rather than aggregating across all groups).

@@

In Sections~\ref{sec:ictstm-example} and \ref{sec:psm-example}, we constructed the disease model using parameter (i.e., \code{params_surv}) objects. Here, we will use a transformed parameter (i.e., \code{tparams_transprobs}) object, which is a very flexible way to construct the transition probability matrices for a cDTSTM. A \code{tparams_transprobs} object requires a 3-dimensional array of transition matrices, where the transition matrices are ordered by PSA sample, treatment strategy, patient, and optionally (in a time-inhomogeneous model) time interval.

@@

Identifiers for the complete array of transitions matrices (inclusive of SOC and the new interventions) are generated with \code{tpmatrix_id()}.

@@

<<rr>>=
x_rr <- as.matrix(transmod_data[, .(soc, new1, new2)])
rr <- cbind(xbeta(x_rr, params_rr_rng$rr_12),
            xbeta(x_rr, params_rr_rng$rr_13))
head(rr) # Corresponds to 1st 6 rows in tpmat_id
@

@@

and the results are summarized. The \code{by_grp} argument of \code{$summarize()} is by default \code{FALSE} so costs and QALYs are aggregated across subgroups here.

@@

\subsubsection{Incremental cost-effectiveness ratio}
The \code{cea_pw()} function computes the ICER. A summary table is produced with \code{icer()} and \code{format()} formats the table for pretty printing. The argument \code{k} is the WTP threshold and is used to compute the INMB. Estimates of incremental QALYs and incremental costs are computed by averaging over PSA samples. By default, 95\% confidence intervals are presented and are computed using quantiles from the PSA.

@@

\subsubsection{Representing decision uncertainty}
A number of common measures can be used to represent decision uncertainty from a CEA, including cost-effectiveness planes, cost-effectiveness acceptability curves (CEAC), cost-effectiveness acceptability frontiers (CEAF), and the expected value of perfect information (EVPI). Plots using \pkg{ggplot2} graphics can be quickly generated with the \code{plot_ceplane()}, \code{plot_ceac()}, \code{plot_ceaf()}, and \code{plot_evpi()} functions, respectively \citep{wickham2016ggplot2}. Output from \code{cea()} and \code{cea_pw()} is readily available, so users may create custom plots as well. Example figures from the code that follows are shown in Figure~\ref{fig:cea-plots}.

@@

One drawback of the CEAC is that the probability of being cost-effective cannot be used to determine the optimal treatment option. Instead, if the objective is to maximize health gain, then decisions should be based on the expected NMB. The CEAF, which plots the probability that the optimal treatment strategy (i.e., the strategy with the highest expected NMB) is cost-effective, is appropriate in this context. A CEAF curve can be created by using the \code{best} column from the \code{mce} element to subset to the treatment strategy with the highest expected NMB for each WTP value.

@@

Alternative approaches might be appropriate in other contexts. For example, if evenly spaced panel data is available, a multinomial logistic regressions can be used to predict transition probabilities. On the other hand, some users may prefer a functional approach that is arguably more human readable. To facilitate this type of analysis, a \code{define_model()} block can be used to construct transition probability matrices using nonstandard evaluation. Each possible combination of the parameters and input data (treatment strategies and patients) is automatically constructed and the user can write custom expressions that create the transition probability matrices as a function of the parameters and input data. The same \code{define_model()} block may also be used to construct utility and cost models, if desired. Examples that leverage multinomial logistic regression and \code{define_model()} are provided on the package website. The latter is used to reproduce two examples (a simple Markov cohort model of an HIV treatment and a time-inhomogeneous Markov cohort model of total hip replacement) from the reference textbook by \citet{briggs2006decision}.

@@

\pkg{mstate} is another package for multi-state modeling that is designed for estimation of non-parametric and semi-parametric (i.e., Cox proportional hazards) models \citep{de2011mstate}. It supports prediction of state probabilities and simulation of multi-state trajectories (to facilitate prediction for clock-reset models) from cumulative hazard functions evaluated at specific times. \citet{williams2017cost} has adapted \pkg{mstate} for cost-effectiveness modeling applications. There are however, some disadvantages of using \pkg{mstate} for health economic modeling since it was not designed for it. \code{mssample()}---the core function for simulating multi-state models---does not natively support heterogeneous target populations, PSA, simultaneous simulation of multiple treatment strategies, or computation of QALYs and costs from the simulated output. It is less computationally efficient than \pkg{hesim} for two main reasons. First, simulation of general cumulative hazards functions is considerably slower than simulation of parametric distributions where efficient random number generators are available. Second, it is written completely in \proglang{R} and incorporating PSA, multiple treatment strategies, or heterogeneous patients results in code that is not vectorized. To compare computational performance, we ran 100 PSA iterations where 1 treatment strategy and 1,000 homogeneous patients were simulated through a 6-state Weibull model during each iteration. The model completed in 0.44 second in \pkg{hesim} and took 34 minutes in \pkg{mstate}, showing that computational considerations are not just an academic concern.

@@

%% -- Conclusion ---------------------------------------------------------------
\section{Conclusion} \label{sec:conclusion}
\pkg{hesim} is a modular and computationally efficient \proglang{R} package for development and analysis of simulation models for economic evaluation. Discrete time cohort and continuous time individual-level models are supported, encompassing Markov (time-homogeneous and time-inhomogeneous) and semi-Markov processes as well as partitioned survival analysis. To facilitate integration of statistical methods and economic models and use of both patient- and aggregate-level data, parameters can be estimated by fitting a statistical model with \proglang{R}, inputting parameter estimates from an external source, or a combination of the two. Simulated costs and QALYs from a PSA can be used for for decision-analysis within a cost-effectiveness framework, allowing users to represent decision uncertainty (e.g., cost-effectiveness planes, cost-effectiveness acceptability curves, and cost-effectiveness acceptability frontiers), and conduct value of information analysis. Algorithms are efficiently coded and written in \proglang{C++} so that approaches that have been considered computationally intensive such as individual-level simulations and PSA can run very quickly. The modular design makes modeling more flexible and it easier to add new features.
